biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-10-03 paper/PaperTRA/PaperTRA.Rmd
wrap paper to TRA
I think the paper is ok for the full paper TRA.
There is a lot of room for improvement, as marked in comments, when submitting to https://www.sciencedirect.com/journal/computers-environment-and-urban-systems/about/call-for-papers#environmentally-and-socially-sustainable-mobility in late november.
@Robinlovelace do you want to take a look at the output, before the submission next week?

Still need to HEAT to be in good heath to make sure that the SE estimates remain the same after the changes.

abstract: | 
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to PT were also estimated and monetized.
  The results indicate that 20% of car trips could switch to the bicycle + PT combination. Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of €230 million. For the PT leg, the transfer from car avoids of at least 8,500 tons of CO~2~eq emissions per year.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
bibliography: bibliography.bib
# Use csl when using default citation_package for Pandoc citeproc
csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
# cls : svproc.csl
# citations with arabic numbers is prefered when using natbib
natbiboptions: numbers

@@

<!-- TODO: -->
<!-- understand where the values from table 2 come from -->
<!-- make sure of the reference cls - try the other cls -->
<!-- see where we can cut to 1 less paragraph -->
<!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to tp transfer Social benefits? maybe to the next paper? -->

@@

Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles. 
For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are also presented, resulting from replacing car trips with cycling.

@@

summary1 = summary1 %>%
  mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
  mutate(Estrategia = paste0(Estrategia*100,"%"))

@@

For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). <!-- NOT CORRECT -->
<!-- place in the end - here is not clear -->
Table \ref{tab:summary21} shows the PT modes with most potential of  in combination with cycling, by mode.
<!-- ???? -->

@@

This scenario unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
<!-- These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips. WE CANNOT SHOW THIS IN THIS PAPER WITH ONLY SCENARIO 3 DATA -->

When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the PT train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.

Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.
<!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity. These estimations used an electricity production life cycle approach. -->
<!-- this co2 needs to be added to the second table and then discuss -->
Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly, with trains offering the greatest potential for substitution (88%).

@@

The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs. <!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper -->
The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment. This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
Improving the safe accessibility to PT interfaces to cyclists and providing bicycle-friendly amenities such as parking facilities can potentially lead to a higher reduction in CO~2~eq emissions, compared to a scenario where individuals shift from car travel to car + PT combination.

@@

The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software to, for instance, understand which potential cycling connections have the highest socio-environmental impacts, in tons of avoided CO~2~eq emissions, or in long term social benefits. <!-- move to introduction? -->

>>>>

abstract: | 
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to PT were also estimated and monetized.
  The results indicate that 10% of trips could be made by bicycle + PT combination. Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of at least €230 million. For the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
bibliography: bibliography.bib
# Use csl when using default citation_package for Pandoc citeproc
csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
# cls : svproc.csl
# citations with arabic numbers is prefered when using natbib
natbiboptions: numbers

@@

<!-- TODO: -->
<!-- see where we can cut to 1 less paragraph -->
<!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to tp transfer Social benefits? maybe to the next paper? -->

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles. 
For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are also presented, resulting from replacing car trips with cycling.

@@

summary1 = summary1 %>%
  mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
  mutate(Estrategia = paste0(Estrategia*100,"%")) |> 
  mutate(TotalTrips = TotalTrips/2,
         Bikebase = Bikebase/2,
         Bikepotential = Bikepotential/2) #this is a quick correction while we donr run biclar entirely again. The error was found and it was adding up the 2 bike legs, instead of making the mean. The SE estimates remain the same because they are from the bikeshare and carshare percentages, that did not change.

@@

For both *direct* and *safe* route profiles, 10% of the daily trips have the potential to be made by a combination of PT and cycling (up to 5 km on bike). This unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
<!-- place in the end - here is not clear -->

Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling.
Trains offer the greatest potential for substitution (88%).
When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.

@@

<!-- These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips. WE CANNOT SHOW THIS IN THIS PAPER WITH ONLY SCENARIO 3 DATA -->

Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.
<!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity. These estimations used an electricity production life cycle approach. -->
<!-- this table could be split into Social benefits and Environmental benefits -->
Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.

@@

The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs. For the PT segment, we did not estimate the social impacts from substituting car trips, although its health benefits would not be as high as shifting to cycling.<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper --> <!--Overall there was very little evidence available on mode switch to public transport. -->

The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment. This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
Improving the safe accessibility to PT interfaces to cyclists and providing bicycle-friendly amenities such as parking facilities can potentially lead to a higher reduction in CO~2~eq emissions, compared to a scenario where individuals shift from car travel to car + PT combination.

@@

The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software. This allows users to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits. <!-- move to introduction? -->

### 2023-10-03 paper/PaperTRA/PaperTRA.tex
wrap paper to TRA
I think the paper is ok for the full paper TRA.
There is a lot of room for improvement, as marked in comments, when submitting to https://www.sciencedirect.com/journal/computers-environment-and-urban-systems/about/call-for-papers#environmentally-and-socially-sustainable-mobility in late november.
@Robinlovelace do you want to take a look at the output, before the submission next week?

Still need to HEAT to be in good heath to make sure that the SE estimates remain the same after the changes.

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
A high proportion of car trips can be replaced by a combination of
public transit and cycling for the first-and-last mile. This paper
estimates the potential for cycling combined with public transit (PT) as
a substitute for car trips in the Lisbon metropolitan area and assesses
its socio-environmental impacts using open data and open source tools. A
decision support tool that facilitates the design and development of a
metropolitan cycling network was developed (\emph{biclaR}). The social
and environmental impacts were assessed using the \emph{HEAT for
Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
car trips to PT were also estimated and monetized. The results indicate
that 20\% of car trips could switch to the bicycle + PT combination.
Shifting to cycling for the first-and-last mile stages can reduce annual
CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
10 years of €230 million. For the PT leg, the transfer from car avoids
of at least 8,500 tons of CO\textsubscript{2}eq emissions per year. This
evidence can support policymakers to prioritize interventions that
reduce the reliance on private motor vehicles.

@@

Table \ref{tab:summary1} presents the LMA total daily trips, the cycling
combined with PT trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are also presented, resulting from
replacing car trips with cycling.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT & Avoided CO2eq (ton/yr) & SE Benefits for 10 years (thousand €)\\
\midrule
4\% & safe & 1 077 028 & 4 624 & 40 770 & 5 920 & 230 270\\
4\% & direct & 1 001 761 & 4 547 & 37 889 & 6 011 & 223 720\\
10\% & safe & 1 077 028 & 4 624 & 104 647 & 15 192 & 591 790\\
10\% & direct & 1 001 761 & 4 547 & 97 218 & 15 414 & 574 200\\
\bottomrule
\end{tabular}
\end{table}

For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
car trips have the potential to be replaced by a combination of PT and
cycling (up to 5 km on bike). Table \ref{tab:summary21} shows the PT
modes with most potential of in combination with cycling, by mode.

@@

This scenario unveils the potential of cycling as a complementary mode
of PT, with the potential to uptake the number of PT trips within the
LMA area by as much as 6.3\% (in addition to the 825 thousand PT trips
reported in the mobility survey).

When comparing the existing PT interfaces (Figure \ref{fig:map1}) with
the bike routes with highest potential to serve as first and last legs
(Figure \ref{fig:map2}) it becomes clear that the PT train interfaces
are the ones that have the highest potential to attract car-to-PT
substituting trips, if their accessibility by bicycle is improved to be
safe.

Table \ref{tab:summary22} presents the results of the avoided emissions
and its monetization for the second leg of the journey, by replacing car
trips with potential TP trips. Regarding the PT segment, the shift from
private car usage would lead to the mitigation of CO\textsubscript{2}
equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
million to €3.5 million yearly, with trains offering the greatest
potential for substitution (88\%).

@@

The social impacts represent 98\% of the socio-environmental benefits
(in value) from replacing car trips to bicycle in first-and-last legs.
The emissions of CO\textsubscript{2}eq that are avoided during both the
initial and final journey segments account for about 75\% of the
emissions avoided during the PT segment. This finding, while expected --
due the zero cycling emissions, should not be overlooked when promoting
the PT use. Improving the safe accessibility to PT interfaces to
cyclists and providing bicycle-friendly amenities such as parking
facilities can potentially lead to a higher reduction in
CO\textsubscript{2}eq emissions, compared to a scenario where
individuals shift from car travel to car + PT combination.

@@

The information available at \emph{biclaR} tool -- an open access
website -- can be downloaded and used with any GIS software to, for
instance, understand which potential cycling connections have the
highest socio-environmental impacts, in tons of avoided
CO\textsubscript{2}eq emissions, or in long term social benefits.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
Económico e Social da Sinistralidade Rodoviária em Portugal}}. {CEGE -
Centro de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
Rodoviária (ANSR)} (2021).}

>>>>

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
A high proportion of car trips can be replaced by a combination of
public transit and cycling for the first-and-last mile. This paper
estimates the potential for cycling combined with public transit (PT) as
a substitute for car trips in the Lisbon metropolitan area and assesses
its socio-environmental impacts using open data and open source tools. A
decision support tool that facilitates the design and development of a
metropolitan cycling network was developed (\emph{biclaR}). The social
and environmental impacts were assessed using the \emph{HEAT for
Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
car trips to PT were also estimated and monetized. The results indicate
that 10\% of trips could be made by bicycle + PT combination. Shifting
to cycling for the first-and-last mile stages can reduce annual
CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
10 years of at least €230 million. For the PT leg, the transfer from car
avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per year.
This evidence can support policymakers to prioritize interventions that
reduce the reliance on private motor vehicles.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles. For the cycling legs of the journey (first and
last legs), the environmental avoided emissions and monetized
socio-environment (SE) benefits are also presented, resulting from
replacing car trips with cycling.

@@

\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario.}
\centering
\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
\toprule
Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT & Avoided CO2eq (ton/yr) & SE Benefits for 10 years (thousand €)\\
\midrule
4\% & safe & 538 514 & 2 312 & 20 385 & 5 920 & 230 270\\
4\% & direct & 500 880 & 2 274 & 18 944 & 6 011 & 223 720\\
10\% & safe & 538 514 & 2 312 & 52 324 & 15 192 & 591 790\\
10\% & direct & 500 880 & 2 274 & 48 609 & 15 414 & 574 200\\
\bottomrule
\end{tabular}
\end{table}

For both \emph{direct} and \emph{safe} route profiles, 10\% of the daily
trips have the potential to be made by a combination of PT and cycling
(up to 5 km on bike). This unveils the potential of cycling as a
complementary mode of PT, with the potential to uptake the number of PT
trips within the LMA area by as much as 6.3\% (in addition to the 825
thousand PT trips reported in the mobility survey).

Table \ref{tab:summary21} shows the potential trips by PT mode to
replace the second leg of the journey, in combination with cycling.
Trains offer the greatest potential for substitution (88\%). When
comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
bike routes with highest potential to serve as first and last legs
(Figure \ref{fig:map2}) it becomes clear that the Train interfaces are
the ones that have the highest potential to attract car-to-PT
substituting trips, if their accessibility by bicycle is improved to be
safe.

@@

Table \ref{tab:summary22} presents the results of the avoided emissions
and its monetization for the second leg of the journey, by replacing car
trips with potential TP trips. Regarding the PT segment, the shift from
private car usage would lead to the mitigation of CO\textsubscript{2}
equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
million to €3.5 million yearly.

@@

The social impacts represent 98\% of the socio-environmental benefits
(in value) from replacing car trips to bicycle in first-and-last legs.
For the PT segment, we did not estimate the social impacts from
substituting car trips, although its health benefits would not be as
high as shifting to cycling.

The emissions of CO\textsubscript{2}eq that are avoided during both the
initial and final journey segments account for about 75\% of the
emissions avoided during the PT segment. This finding, while expected --
due the zero cycling emissions, should not be overlooked when promoting
the PT use. Improving the safe accessibility to PT interfaces to
cyclists and providing bicycle-friendly amenities such as parking
facilities can potentially lead to a higher reduction in
CO\textsubscript{2}eq emissions, compared to a scenario where
individuals shift from car travel to car + PT combination.

@@

The information available at \emph{biclaR} tool -- an open access
website -- can be downloaded and used with any GIS software. This allows
users to, for example, gain insights into which potential cycling
connections have the highest socio-environmental impacts, quantified in
tons of avoided CO\textsubscript{2}eq emissions, or in long term social
benefits.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
Económico e Social da Sinistralidade Rodoviária em Portugal}}. {Centro
de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
Rodoviária (ANSR)} (2021).}
